---
title: "FINC-780: Group Assignment 3 – Trading Rule Back Testing"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In the world of quantitative trading, trend-following strategies have attracted practitioners for decades on the premise that price momentum contains predictive information about future returns. One of the best-known trend-following signals is the death cross, which occurs when a short-term moving average falls below a long-term moving average, commonly interpreted as a bearish signal indicating downward momentum. Its mirror counterpart, the golden cross, occurs when the short-term average rises above the long-term average, signaling bullish momentum. This project explores algorithmic backtesting through the lens of the death-cross and golden-cross trading signal.

The goal here isn't just to see if the rule works, but to ask the "what if" questions. If we had systematically followed the death-cross and golden-cross signal on Tesla (TSLA) over recent years, would we have captured its explosive trends, or would whipsaws and false signals have eroded our returns?

## The Death-Cross / Golden-Cross Signal

The signal is built from two simple moving averages (SMAs). A short-term SMA, which by convention uses a 50-day window, captures recent price momentum, while a long-term SMA, typically 200 days, represents the broader trend. When the 50-day SMA crosses above the 200-day SMA (a golden cross), the strategy enters a long position, betting that upward momentum will persist. When the 50-day SMA crosses below the 200-day SMA (a death cross), the strategy switches to a short position, anticipating continued downside. In this way, the trader always holds a directional position: long when the trend is up, short when the trend is down.

## Link to Behavioral Finance

Why should a moving-average crossover strategy capture real market returns? From a behavioral standpoint, the signal exploits several well-documented investor biases:

**Trend Chasing and Herding:** Investors tend to pile into rising stocks and flee falling ones, creating self-reinforcing trends. The golden-cross signal effectively rides this herding behavior by entering long when momentum is confirmed.

**Underreaction to New Information:** Behavioral research shows that investors are often slow to incorporate new information into prices. Moving averages, by their very construction, detect this gradual price adjustment and signal trend changes only after sustained movement, precisely when underreaction has had time to accumulate.

**Overreaction and Reversals:** When sentiment eventually swings too far, the death cross signals the beginning of a bearish reversal, capturing the period of corrective price action that follows collective overreaction. By systematically switching direction at these crossover points, the strategy attempts to stay aligned with the dominant behavioral regime.

**Anchoring:** Traders and investors anchor to the moving averages themselves. The 50-day and 200-day SMAs are among the most widely watched technical levels, meaning that crossovers can become self-fulfilling prophecies as large numbers of market participants act on the same signal.

# Setup and Libraries

```{r libraries}
library(yfR)
library(TTR)
library(xts)
library(ggplot2)
library(knitr)
library(kableExtra)
```

```{r helper}
# Download daily adjusted prices via yfR and convert to xts
get_prices_xts <- function(ticker, first_date, last_date) {
  df <- yf_get(tickers = ticker,
               first_date = first_date,
               last_date = last_date,
               be_quiet = TRUE)
  price_xts <- xts(df$price_adjusted, order.by = df$ref_date)
  colnames(price_xts) <- "Close"
  return(price_xts)
}
```

# Function 1: Base-Level Back Test

We run the first simulation on TSLA from January 2020 through the end of March 2026 using the standard 50-day and 200-day moving-average windows.

```{r function1}
backtest_ma <- function(ticker, begin, end, short_ma = 50, long_ma = 200,
                        prices_xts = NULL) {

  begin_date <- as.Date(begin, format = "%Y%m%d")
  end_date <- as.Date(end,   format = "%Y%m%d")

  # Download with warm-up buffer, or use pre-loaded xts from Functions 2/3
  if (is.null(prices_xts)) {
    buffer_start <- begin_date - ceiling(long_ma * 1.5)
    close_prices_full <- get_prices_xts(ticker, buffer_start, end_date)
  } else {
    close_prices_full <- prices_xts
  }

  # Compute SMAs on full dataset so values are valid from begin_date
  sma_short_full <- SMA(close_prices_full, n = short_ma)
  sma_long_full <- SMA(close_prices_full, n = long_ma)

  # Subset to the requested test window after SMA computation
  date_range_str <- paste0(begin_date, "/", end_date)
  close_prices <- close_prices_full[date_range_str]
  sma_short <- sma_short_full[date_range_str]
  sma_long <- sma_long_full[date_range_str]

  # Daily arithmetic returns: (P_t - P_{t-1}) / P_{t-1}
  daily_returns <- diff(close_prices) / lag(close_prices, 1)
  colnames(daily_returns) <- "returns"

  # Align returns and SMAs, drop leading NA
  combined <- merge(daily_returns, sma_short, sma_long)
  colnames(combined) <- c("returns", "sma_short", "sma_long")
  combined <- na.omit(combined)

  # Lagged signal avoids look-ahead bias: position at t uses SMAs from t-1
  combined$signal <- ifelse(lag(combined$sma_short, 1) >= lag(combined$sma_long, 1),
                            1, -1)
  combined <- na.omit(combined)

  # Strategy returns
  signal <- combined$signal
  strategy_returns <- combined$returns * signal

  # Count trades (each distinct directional position entered, including initial)
  positions <- as.numeric(coredata(signal))
  pos_changes <- c(positions[1], diff(positions))
  long_trades <- sum(pos_changes ==  2) + ifelse(positions[1] ==  1, 1, 0)
  short_trades <- sum(pos_changes == -2) + ifelse(positions[1] == -1, 1, 0)

  # Percent of time long and short
  n_days <- length(positions)
  pct_long <- sum(positions ==  1) / n_days * 100
  pct_short <- sum(positions == -1) / n_days * 100

  # Cumulative return via geometric compounding
  cum_return <- as.numeric(prod(1 + strategy_returns) - 1) * 100

  # Buy-and-hold benchmark over the same trading days
  bh_return <- as.numeric(prod(1 + combined$returns) - 1) * 100

  summary_df <- data.frame(
    Metric = c("Total Long Trades", "Total Short Trades",
               "Percent Time Long (%)", "Percent Time Short (%)",
               "Cumulative Return (%)", "Buy-and-Hold Return (%)"),
    Value  = round(c(long_trades, short_trades, pct_long, pct_short,
                     cum_return, bh_return), 2)
  )

  results <- list(
    summary = summary_df,
    long_trades = long_trades,
    short_trades = short_trades,
    pct_long = pct_long,
    pct_short = pct_short,
    cum_return = cum_return,
    bh_return = bh_return,
    strategy_ret = strategy_returns
  )

  return(results)
}
```

## Function 1 Results: TSLA, 2020-01-01 to 2026-03-31, SMA(50, 200)

```{r f1_results}
f1 <- backtest_ma("TSLA", "20200101", "20260331", 50, 200)

kable(f1$summary,
      caption = "Death-Cross / Golden-Cross Back Test Summary for TSLA (2020–2026)",
      align = "lr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Interpretation:** The strategy produces a cumulative return of `r round(f1$cum_return, 2)`% over the six-year window, spending `r round(f1$pct_long, 1)`% of the time long and `r round(f1$pct_short, 1)`% short across `r f1$long_trades` long and `r f1$short_trades` short trades. For comparison, a simple buy-and-hold position in TSLA over the same period would have returned `r round(f1$bh_return, 2)`%, making the strategy's underperformance starkly visible. The 200-day SMA is, by construction, a lagging indicator: it takes sustained price movement over many months before the long average shifts enough to trigger a crossover. In Tesla's case, this lag proved damaging during the stock's most violent recoveries. After the COVID crash of early 2020, TSLA surged from its trough;  but the death cross triggered by the initial sell-off kept the strategy short for much of that rally, compounding losses as the short position worked directly against the price appreciation. A similar dynamic played out during the 2023 recovery after the 2022 drawdown. Each day the strategy remained short on a rising stock, it generated a return equal to the negative of that day's gain. These extended periods of "short during a bull run" account for the gap between the strategy and the buy-and-hold benchmark. Additionally, because Tesla exhibits a strong upward trend over this period, a strategy that frequently takes short positions is structurally disadvantaged relative to buy-and-hold, further contributing to the observed underperformance. The trade count reflects how many full crossover events occurred — each one representing a complete directional reversal that would also incur transaction costs in practice, further eroding returns.

# Function 2: Simulate Multiple Back-Test Periods

It's easy to get a "lucky" or "unlucky" result by picking one specific window. To assess whether the death-cross signal is robust, we break the 2018–2026 period into rolling 3-year sub-periods. Each sub-period spans exactly three calendar years (e.g., 2018–2020, 2019–2021, ..., 2024–2026). To ensure valid SMA values from the first day of each sub-period, the initial data download includes a one-year buffer before the overall start year.

```{r function2}
backtest_multi_period <- function(ticker, test_years, date_range,
                                  short_ma = 50, long_ma = 200) {

  start_year <- as.numeric(date_range[1])
  end_year <- as.numeric(date_range[2])

  # Download with one-year buffer for SMA warm-up
  buffer_start <- as.Date(paste0(start_year - 1, "-01-01"))
  all_prices <- get_prices_xts(ticker,
                                 buffer_start,
                                 as.Date(paste0(end_year, "-12-31")))

  # Roll test_years-year windows; last valid start = end_year - test_years + 1
  results_list <- list()
  period_labels <- c()
  cum_returns <- c()

  for (yr in start_year:(end_year - test_years + 1)) {
    begin <- paste0(yr, "0101")
    end <- paste0(yr + test_years - 1, "1231")
    label <- paste0(yr, "-", yr + test_years - 1)

    res <- backtest_ma(ticker, begin, end, short_ma, long_ma,
                       prices_xts = all_prices)

    results_list[[label]] <- res
    period_labels <- c(period_labels, label)
    cum_returns <- c(cum_returns, res$cum_return)
  }

  # Compute means across all iterations
  mean_long_trades <- mean(sapply(results_list, function(x) x$long_trades))
  mean_short_trades <- mean(sapply(results_list, function(x) x$short_trades))
  mean_pct_long <- mean(sapply(results_list, function(x) x$pct_long))
  mean_pct_short <- mean(sapply(results_list, function(x) x$pct_short))
  mean_cum_return <- mean(sapply(results_list, function(x) x$cum_return))

  summary_df <- data.frame(
    Metric = c("Mean Long Trades", "Mean Short Trades",
               "Mean Percent Time Long (%)", "Mean Percent Time Short (%)",
               "Mean Cumulative Return (%)"),
    Value  = round(c(mean_long_trades, mean_short_trades,
                     mean_pct_long, mean_pct_short, mean_cum_return), 2)
  )

  # Bar chart of cumulative return by sub-period
  plot_df <- data.frame(Period = period_labels, Cumulative_Return = cum_returns)
  plot_df$Period <- factor(plot_df$Period, levels = period_labels)

  p <- ggplot(plot_df, aes(x = Period, y = Cumulative_Return, fill = Period)) +
    geom_bar(stat = "identity", width = 0.6) +
    geom_text(aes(label = paste0(round(Cumulative_Return, 1), "%")),
              vjust = -0.5, size = 3.5) +
    labs(title = paste("Cumulative Return by Sub-Period:", ticker),
         subtitle = paste(test_years, "-Year Testing Windows  |  SMA(",
                          short_ma, ",", long_ma, ")", sep = ""),
         x = "Period", y = "Cumulative Return (%)") +
    theme_minimal() +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold"))

  return(list(summary = summary_df, plot = p, details = results_list))
}
```

## Function 2 Results: TSLA, 3-Year Windows, 2018–2026, SMA(50, 200)

```{r f2_results, fig.width=8, fig.height=5}
f2 <- backtest_multi_period("TSLA", 3, c("2018", "2026"), 50, 200)

kable(f2$summary,
      caption = "Mean Results Across 3-Year Sub-Periods for TSLA",
      align = "lr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

print(f2$plot)
```

**Interpretation:** The bar chart shows cumulative returns across seven rolling 3-year windows (2018–2020 through 2024–2026), each representing a distinct market regime in Tesla's history. The mean cumulative return of `r round(f2$summary$Value[5], 2)`% provides a single summary of average profitability across all windows. Considerable variation across sub-periods is the expected and informative result: windows that coincide with Tesla's explosive directional runs (upward or downward) may show large positive returns if the strategy happened to be on the correct side, while windows that include major V-shaped recoveries — where the 200-day SMA's lag forces the strategy to remain short through an aggressive rebound — are likely to show the worst losses. A strategy whose profitability varies across sub-periods is inherently regime-dependent, which is expected for trend-following approaches. However, the magnitude of variation observed here raises concerns about consistency and predictability, as performance depends heavily on the specific market environment. The mean return and dispersion across windows together characterize whether the death-cross signal represents a durable edge or a regime-specific outcome.

# Function 3: Simulate Multiple Short Moving-Average Windows

Finally, we want to see how sensitive the strategy is to the choice of the short moving-average window. Holding the long moving average fixed at 200 days, we vary the short window from 20 to 80 days in increments of 20.

```{r function3}
backtest_multi_short_ma <- function(ticker, begin, end, low_short, high_short,
                                     ma_increment, long_ma = 200) {

  begin_date <- as.Date(begin, format = "%Y%m%d")
  end_date <- as.Date(end,   format = "%Y%m%d")

  # Download once with warm-up buffer, reused across all short MA iterations
  buffer_start <- begin_date - ceiling(long_ma * 1.5)
  all_prices <- get_prices_xts(ticker, buffer_start, end_date)

  short_windows <- seq(low_short, high_short, by = ma_increment)
  long_trades <- numeric(length(short_windows))
  short_trades <- numeric(length(short_windows))
  cum_returns <- numeric(length(short_windows))

  for (i in seq_along(short_windows)) {
    current_short <- short_windows[i]
    res <- backtest_ma(ticker, begin, end, current_short, long_ma,
                       prices_xts = all_prices)
    long_trades[i] <- res$long_trades
    short_trades[i] <- res$short_trades
    cum_returns[i] <- res$cum_return
  }

  results_df <- data.frame(
    Short_MA_Window = short_windows,
    Long_Trades = long_trades,
    Short_Trades = short_trades,
    Cumulative_Return = round(cum_returns, 2)
  )

  # Line plot of cumulative return vs short MA window
  p <- ggplot(results_df, aes(x = Short_MA_Window, y = Cumulative_Return)) +
    geom_line(color = "steelblue", linewidth = 1) +
    geom_point(color = "steelblue", size = 2.5) +
    geom_text(aes(label = paste0(round(Cumulative_Return, 1), "%")),
              vjust = -1, size = 3.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = paste("Cumulative Return vs. Short MA Window:", ticker),
         subtitle = paste("Long MA fixed at", long_ma,
                          "days  |  Back Test:", begin, "to", end),
         x = "Short Moving-Average Window (days)",
         y = "Cumulative Return (%)") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))

  return(list(table = results_df, plot = p))
}
```

## Function 3 Results: TSLA, 2020–2026, Short MA 20–80 (step 20), Long MA 200

```{r f3_results, fig.width=8, fig.height=5}
f3 <- backtest_multi_short_ma("TSLA", "20200101", "20260331", 20, 80, 20, 200)

kable(f3$table,
      caption = "Strategy Performance Across Short MA Windows (Long MA = 200)",
      row.names = FALSE, align = "cccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

print(f3$plot)
```

**Interpretation:** The table and plot reveal how sensitive the death-cross / golden-cross strategy is to the choice of the short moving-average window. A shorter window (20 days) makes the strategy more reactive, generating more frequent crossover signals and therefore more total trades. While this increases the potential to capture short-lived trends, it also amplifies exposure to whipsaw losses — situations where the averages cross back and forth without establishing a durable trend, triggering a series of poorly-timed reversals. A longer short window (80 days) smooths out noise but introduces meaningful additional lag on top of the already-slow 200-day average, causing the strategy to enter and exit positions well after a genuine trend has developed. The trade count column illustrates this trade-off directly: shorter windows generate many more signals, while longer windows converge toward the minimal-signal behavior of the standard 50-day convention. The results show that performance improves significantly as the short moving-average window increases, with longer windows (60–80 days) producing positive cumulative returns while shorter windows remain negative. This indicates that the strategy benefits from reduced trading frequency and noise filtering, rather than the underperformance being purely a structural feature across all parameter choices.

# Discussion

After examining the results from all three functions, several key findings emerge.

**Function 1** provides the baseline result for the standard 50/200-day crossover on Tesla over a six-year window. The strategy's cumulative return of `r round(f1$cum_return, 2)`% compares poorly against a buy-and-hold return of `r round(f1$bh_return, 2)`% over the same period — a gap of `r round(f1$bh_return - f1$cum_return, 2)` percentage points. This is best understood not as a failure of the death-cross concept generally, but as an illustration of the strategy's specific vulnerability on high-volatility momentum stocks. The 200-day SMA's inherent lag — which is both its feature and its flaw — meant that the strategy frequently found itself short precisely during Tesla's most aggressive upward moves. Because the strategy is always in a directional position (never in cash), every day spent short on a rising stock compounds losses at the same rate the buy-and-hold investor was accumulating gains. Transaction costs are not included in this figure; each crossover represents a full position reversal that would incur bid-ask spreads and brokerage commissions in practice, reducing the reported cumulative return further.

**Function 2** demonstrates the importance of not relying on a single time window. By rolling 3-year sub-periods across 2018–2026, we can assess whether the strategy performs consistently or is highly dependent on the specific market regime. Tesla's price history includes extreme momentum phases (the 2020 rally), extended drawdowns (2022), and range-bound consolidation, each of which presents different challenges for a trend-following signal. A strategy that earns positive returns in some periods and large negative returns in others is regime-dependent and not reliably profitable,  a critical consideration before deploying real capital.

**Function 3** addresses parameter sensitivity. The choice of 50 days for the short moving average is a widely adopted convention, not an optimization. By sweeping across 20, 40, 60, and 80 days, we can assess whether the strategy's performance is robust to this design choice. The variation in cumulative returns across short windows shows that parameter choice materially affects performance, with longer windows producing stronger results. This suggests that the strategy's effectiveness is sensitive to design choices rather than being purely a structural limitation of the crossover approach.

From a behavioral finance perspective, the death-cross and golden-cross signals are interesting precisely because they are so widely followed. When a large proportion of market participants watch the same 50-day and 200-day moving averages, crossover events can become self-fulfilling: a golden cross triggers buying, which pushes prices higher, which reinforces the signal. This herding dynamic is distinct from the signal capturing fundamental value — it captures a coordination mechanism among technical traders. This distinction matters for interpreting the results: a strategy that profits from investor behavior rather than economic fundamentals is inherently fragile, and its effectiveness depends on the continued participation of traders acting on the same indicators.

On the question of economic meaningfulness: the strategy's cumulative return of `r round(f1$cum_return, 2)`% versus a buy-and-hold return of `r round(f1$bh_return, 2)`% represents deeply negative alpha. Accounting for turnover and execution lag would only worsen that figure. Each crossover involves a full position reversal, meaning slippage on both the close and the open. For a stock as liquid as Tesla, per-trade costs are modest, but they are always additive losses on top of an already-negative strategy return. Lagging execution — where the signal is confirmed only after the market close and the trade is executed at the following day's open — introduces additional gap risk that is not captured in this backtest. A more complete evaluation would also consider risk-adjusted performance metrics such as the Sharpe ratio or volatility of returns, as a strategy that underperforms in absolute terms may still offer diversification or downside protection benefits under certain conditions.

# Conclusion

This backtest demonstrates how a widely followed technical indicator, the 50/200-day moving-average crossover, performs when applied systematically to Tesla. The death-cross and golden-cross signals capture the essence of trend-following: ride momentum when it confirms, and reverse when the trend breaks. The three functions we built provide a complete analytical toolkit, examining a single base case, testing robustness across time periods, and assessing parameter sensitivity.

The key takeaway is that the death-cross strategy fails on TSLA over this period not because trend-following is conceptually unsound, but because the 200-day SMA's lag is too slow for a stock that recovers violently and quickly from its drawdowns. The strategy's always-invested nature, it is never in cash — means that being on the wrong side of a major move is particularly costly. Moving forward, potential improvements could include adding a confirmation filter (requiring the crossover to persist for several days before acting), incorporating a cash or exit option during ambiguous regimes, using exponential rather than simple moving averages to reduce lag, or combining the crossover signal with volatility-based position sizing to limit downside during uncertain market conditions.

Beyond the results themselves, this assignment reinforced several lessons about rigorous quantitative analysis. First, backtesting is deceptively easy to do incorrectly: subtle mistakes like computing moving averages only over the test window (rather than a buffered history) or failing to lag the signal can introduce look-ahead bias that makes a strategy appear far better than it truly is. Second, a single backtest window can be deeply misleading — the rolling sub-period analysis in Function 2 made clear that the same strategy can look like a winner or a loser depending entirely on which years are examined, which is a warning against cherry-picking results. Third, the buy-and-hold benchmark reframes the entire discussion: a strategy does not need to lose money to be a bad strategy; it only needs to underperform the passive alternative after accounting for its costs and risks. Finally, connecting the mechanics of the signal to behavioral finance concepts — anchoring, herding, underreaction — gave us a framework for thinking about *why* such rules might work in some environments and fail in others, rather than treating the backtest result as a black box. These lessons apply well beyond the death-cross signal to quantitative research more broadly.

# References

- Murphy, J. J. (1999). *Technical Analysis of the Financial Markets*. New York Institute of Finance.
- Brock, W., Lakonishok, J., & LeBaron, B. (1992). Simple Technical Trading Rules and the Stochastic Properties of Stock Returns. *Journal of Finance*, 47(5), 1731–1764.
- Shiller, R. J. (2000). *Irrational Exuberance*. Princeton University Press.
- De Bondt, W. F. M., & Thaler, R. (1985). Does the Stock Market Overreact? *Journal of Finance*, 40(3), 793–805.
- TTR: Technical Trading Rules. R Package. CRAN.
